---
title: "Numerical Descriptive Statistics"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered

tutorial:
  id: "numerical-descriptives"
  version: 0.5
---
```{r setup, include=FALSE}
library(ggplot2)
library(learnr)
library(gradethis)
library(datasets)
data(mtcars)
data("iris")
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
learnr::tutorial_options(exercise.timelimit = 60)
```

## Descriptive Statistics

###

One of the first steps that you might take once you have some data in R is to look at descriptive statistics. Remember, the types of descriptive statistics that you use will be different depending on whether it is categorical or numerical data. We will look at numerical descriptive statistics and graphs here; check out the video tutorial in order to learn about how to do descriptive statistics and graphs for categorical data.
![categorical descriptives](https://www.youtube.com/watch?v=z4l1Ff97gHo&t=38s)

### Example Dataset: Iris

In the examples for this tutorial, we will use the `iris` dataset provided in `datasets` package in R.


```{r}
head(iris)
```

This is a dataset about measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica. You can find information about the dataset [here](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris).


## Measures of Center and Spread

###

Many of the functions for getting numerical summaries of center and spread are intuitive in R. 

|Function|Type of Plot|
|---|---|
|`mean(x)`|Finds the mean of a vector `x`|
| `median(x)` | Finds the median of a vector `x`.
| `sd(x)` | Finds the standard deviation of a vector `x` |
| `variance(x)` | Finds the variance of a vector `x` |
| `IQR(x)` | Finds the IQR of a vector `x`|
| `summary(x)` | Finds the minimum, 1st quartile, median, mean, 3rd quartile and maximum of a vector `x`, or of each column in a data frame `x`|

In general, these are used on a vector, rather than a data frame, so you'll need to be careful to specify the correct column within the data frame. Remember that you can use `$` notation to extract a specific column from the data frame.


```{r, echo=TRUE}
mean(iris$Sepal.Length)
sd(iris$Sepal.Length)
```

### Practice Problem

How would you find the median `sepal length` of the iris flowers?

```{r practice1, exercise=TRUE}
```

```{r practice1-check}
grade_result(
  pass_if(~ identical(.result, median(iris$Sepal.Length)), "Good Job!"),
  fail_if(~ TRUE)
)
```

## Plotting with base R

### Visualizing Numerical Variables

The basic functions for plotting numerical variables in `R` are shown in the table below. 

|Function|Type of Plot|
|---|---|
|`hist(x)`|histogram|
| `boxplot(x)` | boxplot |
| `plot(x,y)` | scatterplot |

For example, to get a simple histogram, we can use the following command.

```{r}
hist(iris$Sepal.Length) 
```

We can make adjustments to this graph by adding additional **arguments** to the `hist` function. Arguments are the inputs that we provide a function. Above, we only provided one argument: the data to be plotted. We can also specify that we want a different number of bins, for example.

```{r}
hist(iris$Sepal.Length, breaks = 10) # The "breaks = 10" argument sets the number of bins
```

Now, suppose we want to use the `relative frequency` of the flowers to plot the histogram instead of absolute `frequency` in the y-axis. We can provide an argument `probability = TRUE` inside the `hist` function for the purpose. In that case, the total area under the histogram bins will sum up to $1$.

```{r}
hist(iris$Sepal.Length,probability = TRUE)
```

We can redefine the names of the axes, or even the name of the plot using some specific arguments like `xlab`, `ylab` and `main`. But the names should be provided as a `string` i.e. be careful to put "" around the names! For example,

```{r}
hist(iris$Sepal.Length, breaks = 10, probability = TRUE, xlab = "Sepal Length",ylab = "Relative Frequency", main ="Histogram of Iris Data")
```

You can play around with the function `hist()` to discover utilities of different arguments! Check the documentation [here](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/hist).

### Boxplots

We can make boxplots using the `boxplot` function.

```{r}
boxplot(iris$Sepal.Length)
```

Similar to the case of `hist()` function, we can add a title here too. Why not try your hand to put a name to the boxplot?

### Practice Problems

Create a boxplot of `Petal.Length` of the iris flowers and add a title **Boxplot of Petal Length** to the plot.
Submit the required code below.

```{r practice2, exercise = TRUE}
```


```{r practice2-solution}
boxplot(iris$Petal.Length,main = "Boxplot of Petal Length")
```
 
```{r practice2-code-check}
grade_code()
```

## Plotting with `qplot`

### 

As we mentioned before in the `intro-r` tutorial, there are external packages that contain various functions, data, and other tools to help us do more with R. One of these is the `ggplot2` package, which contains many functions for generating visually pleasing graphs.

```{r}
# Run install.packages if you do not have it installed already.
# install.packages('ggplot2')
library(ggplot2)
```

The `ggplot2` package is incredibly rich and highly customizable for an experienced user. There is an [entire book](https://ggplot2-book.org/) you can read through on using `ggplot2` to create graphics, and the code can get quite complicated. We won't go over all of that in this short introduction to producing graphs. Rather, in this tutorial, we'll focus on one function, `qplot`, which is a quick way of getting nice looking graphs without doing much work. 

### Using `qplot`

The idea behind `qplot` is that it can be used to generate graphs very quickly using just one function. To do this, it tries to detect the type of data being plotted, and generates an appropriate graph based on this. Let's try using `qplot` with sepal length of iris flowers. 

```{r, echo = TRUE}
qplot(iris$Sepal.Length)
```

###

Instead of using `breaks` as an argument to specify the number of bins we want for the histogram, we use `bins` with `qplot`. In addition, we can use the `data` argument to make it easier to read and parse what is happening in the function. We show two ways of doing the same thing below, with the first commented out.

```{r, echo = TRUE}
# This is the same thing
# qplot(iris$Sepal.Length, bins = 10)
qplot(Sepal.Length, data = iris, bins = 10)
```

### 

This made a histogram, exactly like we might have wanted. If we wanted a boxplot instead, we can add an additional argument to specify this. Note that we specify `y=` because we want to plot the graph on the y-axis. The `qplot` function for boxplots only works for putting the continuous variable on the y-axis -- in the categorical-descriptives analysis, we can have side-by-side boxplots which can show the distributions by group in a categorical variable, and we'll have to specify `x=` there. For now, just remember to specify `y=` when using `qplot` for boxplots.

```{r, echo = TRUE}
qplot(y = Sepal.Length, data = iris, geom = 'boxplot')
```

If we want to make a scatterplot, we can use it very similar to how we would use `plot`. 

```{r, echo = TRUE}
# This does the same thing
# qplot(iris_scores$Sepal.Length, iris$Sepal.Width)
qplot(Sepal.Length, Sepal.Width, data = iris)
```

### Adding a title

We can add a title like we did before using the `main` argument.

```{r, echo = TRUE}
qplot(Sepal.Length, Sepal.Width, data = iris, main = "Scatterplot of sepal length and sepal width of Iris flowers")
```

## Exercises

Consider the `mtcars` dataset in package `datasets`.

```{r}
data(mtcars)
head(mtcars)
```
Check the documentation for the dataset [here](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars), if needed.


### Exercise 1

What is the overall mean of number of cylinders used in the cars? 

```{r exercise1, exercise=TRUE}
```

```{r exercise1-check}
grade_result(
  pass_if(~ identical(.result, mean(mtcars$cyl)), "Good Job!"),
  fail_if(~ TRUE)
)
```


### Exercise 2

What is the mean number of cylinders for cars that have a V-shaped engine? 

```{r exercise2, exercise=TRUE}
```

```{r exercise2-check}
grade_result(
  pass_if(~ identical(.result, mean(mtcars$cyl[mtcars$vs == 0])), "Good Job!"),
  fail_if(~ TRUE)
)
```

### Exercise 3

Create a histogram of the `Miles/gallon` for all cars who had a V-shaped engine. Use 15 bins in the histogram, and use the base R method (not using `qplot`). 

```{r exercise3, exercise=TRUE}
```

<div id="exercise3-hint">
**Hint:** This is a two step process. First, you need to subset, then you need to create the graph.
</div>

```{r exercise3-check}
grade_result(
  pass_if(~ identical(.result, hist(mtcars[mtcars$vs == 0,'mpg'], breaks = 15)), "Good Job!"),
  pass_if(~ identical(.result, hist(mtcars$mpg[mtcars$vs == 0], breaks = 15)), "Good Job!"),
  # pass_if(~ isTRUE(all.equal(.result, qplot(mtcars[mtcars$vs == 0,'mpg'], bins = 15))), "Good Job!"),
  fail_if(~ TRUE)
)
```


### Submitting work

```{r submission-setup}
submission_code <- function(UID){
  httr::sha1_hash('numdesc',as.character(UID))
}
```

Generate your submission code by putting in your UID in the function below. For example, if your UID is `2`, then your code should look like `submission_code(UID = 2)`
```{r submission, exercise = TRUE}
# Replace the number below with your UID
submission_code(2)
```


