---
title: "Data Cleaning in R"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "R-data-cleaning"
  version: 0.5
---

```{r setup, include=FALSE}
library(MASS)
library(learnr)
library(gradethis)
library(readxl)
library(xlsx)
library(datasets)
library(psych)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

## Overview

This tutorial mainly focuses on an overall summary of what you learnt so far and give you a brief idea about how to handle missing observations in your data. Let's begin!

## Review of Earlier Lessons

- We will start by first opening the XLSX file in the dataset. 
- Remember to create a original file and a working file.
- Remember to load the package `readxl` in your local environment before reading a .xlsx file.

```{r}
data.original1 <- read_xlsx("D:/BSOS_R_Tutorial/PSYC_Tutorials/data_psych/STAT200_Data_Update_4.8_original.xlsx")
data.working <- data.original1
```


## Structure & Summary of Dataset

```{r}
#str() allows us to see the structure of the variables in the dataset
str(data.working)

#summary provides us with a summary of the variables in the dataset
summary(data.working)
```

## Describe / Describe.By() Datasets


### install/library

- We first begin with the regular install/library process
- Note if you have it already installed in the computer, you do not need to install the package

```{r}
#install.packages(psych)
#library(psych)
```

### describe()

```{r}
#describe() also provides us a clean summary/information on the variables in the dataset
describe(data.working)
  #describe(dataset)
```

*Note: You can find more on `describe` function [here](https://www.rdocumentation.org/packages/psych/versions/2.0.12/topics/describe)*.

### describe.by()

```{r}
#describe.by() provides us a clean summary/information on the variables in the dataset FOR A SPECIFIC GROUP

#Preparing the Data into Factors

#Original Dataset Structure
str(data.original1)

#Converting Stat Anxiety from Numeric to Factors
data.working$Stat.Anxiety_Name <- as.factor(data.working$Stat.Anxiety_Name)
  #Low Anxiety
  #Medium Anxiety
  #High Anxiety

#Observing Change from Numeric to Factor
str(data.working)

#Running the Code
describe.by(data.working, data.working$Stat.Anxiety_Name)
  #describe(dataset, dataset$group)
```

*Note: For more information on `describe.by()`, refer to this [link](https://www.rdocumentation.org/packages/psych/versions/2.0.12/topics/describeBy)*.


## Review of Subsetting in R

In this section, we quickly review some of the things about subsetting that you learnt in the last tutorial.

### Selecting specific Variables in the Dataset

```{r}
#This is how to Extract a specific variable via $
data.working$Sleep
data.working$Motivation
  #dataset$variable

#This is how to Extract a specific variables via []
data.working[,2]
data.working[,3]
data.working[,2:3]
  #dataset[row,column]

#This is how to Extract a specific row via []
data.working[1,]
data.working[3,]
data.working[1:3,]

#This is how to Extract a specific variable and row via []
data.working[1,1] #row 1 and column 1
data.working[1:3, 1:3] #row 1 to 3 & column 1 to 3
```

Next, we look at exam scores by each factor of Statistics Anxiety by directly referencing variables we are interested in using [ ] and logical operators.

### Factors

- Here we subset for Factors.

```{r}
#Exam Scores for Students with Low Stat Anxiety (Stat Anxiety == 1)
data.working$Scores[data.working$Stat.Anxiety  == 1]

#Exam Scores for Students with High Stat Anxiety (Stat Anxiety == 3)
data.working$Scores[data.working$Stat.Anxiety == 3]
```

### Numeric

- Here we subset for numerical information.

```{r}
#Exam Scores for Students with Motivation Equal to 5
data.working$Scores[data.working$Motivation  == 5]

#Exam Scores for Students with Motivation Greater than 5
data.working$Scores[data.working$Motivation > 5]

#Exam Scores for Students with Motivation Less than 5
data.working$Scores[data.working$Motivation < 5]
```


### AND Logical Operator &

```{r}
#Exam Scores for Statistics Anxiety Low (1) & Motivation Greater than 5
data.working$Scores[data.working$Stat.Anxiety == 1 & data.working$Motivation > 5]

data.working$Scores[data.working$Stat.Anxiety == 1 & data.working$Sleep > 300]


#Exam Scores for Statistics Anxiety High (3) & Motivation Less than or Equal to 5
data.working$Scores[data.working$Stat.Anxiety == 3 & data.working$Motivation <= 5]
```

## Cleaning Datasets

Now we move towards something new and interesting, which is data cleaning using R!
Much of work in RStudio and data analysis involves processing datasets. This is not the glamorous part of statistical analysis but it makes up more than 75% of the job itself. In this section, we will discuss three popular methods for data cleaning.

First, let's check the Total Sample Size of Dataset.

```{r}
#Sample Size Original/Working Datasets
nrow(data.original1)
nrow(data.working)

#N = 90 as we have 90 participants in total
```

Now, lets reference the variables that have missing data to them. Here when we reference the missing_data variable, we can see that we are missing some values below.

```{r}
data.working$Stat.Anxiety
data.working$Stat.Anxiety.Missing

#Note: The Missing Data is referenced as NA
```

**The missing data is very common and may be due to:**
    
- Participant Fatigue
- Privacy Concerns
- Connectivity Issues
   

For data analysis, the easiest way to address missing values consist of three methods:

- Listwise Deletion
- Casewise Deletion
- Mean/Median Imputation

Now, we will discuss these three methods.

## Listwise Deletion

- This method refers to deleting all rows in the dataset that contain at least one missing data value in all of the variables.
- `na.omit()` can be used to limit data to only rows with data for all variables.
- One Problem with this method is that it can eliminate a lot of data.

```{r}
data.listwise.deletion <- na.omit(data.working)
data.listwise.deletion #Notice the Missing_Data Column

#Compare Listwise & Working
head(data.original1$Stat.Anxiety.Missing)
head(data.listwise.deletion$Stat.Anxiety.Missing)

nrow(data.original1) #N = 90
nrow(data.listwise.deletion) #N = 67
```

## Casewise Deletion

- This method refers to just ignoring the data for a certain type of analysis when it is missing.
- R requires us to do this for many analyses by ignoring the data that is missing a value; but also does it automatically for some analyses by ignoring the data that is missing a value.
    
### Example where R Does not do it Automatically

- To specify casewise deletion manually, you can input a `na.rm = TRUE` within the function itself.

```{r}
#R Does not Work
mean(data.working$Stat.Anxiety.Missing)
median(data.working$Stat.Anxiety.Missing)

#R Works when we do it Manually
mean(data.working$Stat.Anxiety.Missing, na.rm = TRUE)
median(data.working$Stat.Anxiety.Missing, na.rm = TRUE)
```

### Example where R Does it Automatically

- However, sometimes R functions do casewise deletion automatically!
- The `cor.test()` function does this!
- Unfortunately, we have to check each function to see how it treats missing data. 

### Example where R does Casewise Deletion Automatically

```{r}
#Correlation between Statistics Anxiety (Missing) & Exam Scores
cor.test(data.working$Stat.Anxiety.Missing, data.working$Scores)
```

## Median/Mean Imputation

- This method refers to filling in the missing data values with the mean or median of said variable.
- For instance if the mean is 5 for motivation variable, we would fill in all of the `NA`s for motivation variable with 5's.
- The benefits of mean/median imputation is that it allows us to have more data to use for our analyses.
- The major downside is that it reduces our variability in the dataset.

    
### Preparing Data for Mean/Median Imputation

```{r}
#Create a Copy for Mean/Median Imputation
mean_impute <- data.working
median_impute <- data.working

#Calculate Mean/Median for Stat Anxiety
mean(mean_impute$Stat.Anxiety.Missing, na.rm = TRUE)
median(mean_impute$Stat.Anxiety.Missing, na.rm = TRUE)
```

### Doing Mean Imputation

- The general format of writing the code is:

"`r #Name_of_Dataset$Variable_to_Reference[is.na(#name_of_Dataset$Variable_to_Reference)] <- Value_to_Replace_MissingData`"

```{r}
#Mean Imputation
mean_impute$Stat.Anxiety.Missing[is.na(mean_impute$Stat.Anxiety.Missing)] <- 1.9868642

#Observing Change
mean_impute$Stat.Anxiety.Missing

#Compare to Old
data.working$Stat.Anxiety.Missing
```

*Note: Significant figures also now change since we added a bunch of figures when we added the mean (e.g., 1.9868642)!*

### Doing Median Imputation

- Same format as Mean Imputation, except now we just use 2 instead of 1.9868642 because the median of the observed data was 2.
    
```{r}
#Median Imputation
median_impute$Stat.Anxiety.Missing[is.na(median_impute$Stat.Anxiety.Missing)] <- 2

#Observing Change
median_impute$Stat.Anxiety.Missing

#Compare to Old
data.working$Stat.Anxiety.Missing
```

*Note: Significant figures do not change since we did not add extra figures after adding the median (e.g, 2)!*

Time for some fun exercises!

## Exercises

###

Consider the `airquality` data set here again.

```{r echo=TRUE, eval=TRUE}
data("airquality")
```

### Exercise 1

How will you find the mean of `Ozone` concentration, given that you are supposed to delete any missing value for this variable?

```{r exercise1, exercise=TRUE, echo=FALSE}
```

```{r exercise1-check, echo=FALSE}
grade_result(
  pass_if(~ identical(.result, mean(airquality$Ozone,na.rm = TRUE)), "Good Job!"),
  fail_if(~ TRUE)
)
```

### Exercise 2

How would you impute the missing values for `Ozone` with their median?

```{r exercise2, exercise=TRUE, echo=FALSE}
```

```{r exercise2-solution, echo=FALSE}
airquality$Ozone[is.na(airquality$Ozone)] <- median(airquality$Ozone,na.rm = TRUE)
```

```{r exercise2-check, echo=FALSE}
grade_code("Excellent")
```

###

### Exercise 3

How would you omit all rows containing missing values?

```{r exercise3, exercise=TRUE, echo=FALSE}
```

```{r exercise3-check, echo=FALSE}
grade_result(
  pass_if(~ identical(.result, na.omit(airquality)), "Well done!"),
  fail_if(~ TRUE)
)
```

###

### Submitting Work

```{r submission-setup, echo=FALSE}
submission_code <- function(UID){
  httr::sha1_hash('data_cleaning',as.character(UID))
}
```

Generate your submission code by putting in your UID in the function below. For example, if your UID is `2`, then your code should look like `submission_code(UID = 2)`

```{r submission, exercise = TRUE, echo=FALSE}
# Replace the number below with your UID
submission_code(2)
```

##



