---
title: "Data sets in R"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "R-functions"
  version: 0.5
---

```{r setup, include=FALSE}
library(MASS)
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

# What to learn here?

Welcome to another lesson in R. In this lesson, we will talk about accessing datasets from different packages inside R as well as from outside R. Then we touch upon some useful functions that can be used on those datasets to extract interesting information. Let's begin!

## Accessing Datasets from Rstudio Packages

- At first we will learn how to install any package in R. Here we are going to install the package `datasets`.

```{r echo =TRUE, eval= FALSE}
install.packages("datasets")
```

- Keep in mind that once you have installed any package, either remove this line from your program or add a # before it. Otherwise, if you knit the Markdown file with the line as it is again, R studio might run into problem while trying to install it, because the package is already installed.

- Next, we need to bring/call the package into our working environment to use any function from that package. Here we show how to do that.

```{r echo=TRUE}
library(datasets)
```


### ChickWeight & VADeaths Dataset
We will access these two datasets below in this package

    - ChickWeight = Weight versus age of baby chickens on different diets
    - VADeaths = Dataset of Death Rates in Virginia (1940)
    
### Good Practice/General Recommendation for Loading Datasets into R

- It is always useful to load the data set and store it into an object in R.
- That enables us to refer to that data set anytime we want.

```{r echo=TRUE}
#Original Dataset
ChickWeight
VADeaths

#Dataset we work with
dat.working1 <- ChickWeight
dat.working2 <- VADeaths

#Viewing Dataset
dat.working1
dat.working2
```

### Try it!

1. Try to write the code for the following task.
- Load the package "MASS".
- Load the data set `survey` and save it as `dat.survey` in R.

```{r code_sandbox, exercise=TRUE, exercise.cap="datasets"}
```

Alright! Let's move on to the part of the lesson where we see how to load data sets from our local computers into R.

## Opening Datasets from Local Files

Most of the time in real life, you will work with data that have been stored by some means into an external address, rather than inside any R package. You have to bring that data set into R before performing any kind of analysis, right? Here is the way how you do it.

### Installing Packages to Load Datasets

- To open up excel files and csv files we first need to install and then open a package that allows us to `read` the files into R.
- Below are popular packages for opening excel & Csv files.

    
```{r echo=TRUE}
#Excel Datasets
#install.packages("readxl")
library(readxl)

#CSV Datasets
#install.packages("readr")
library(readr)


#Note: Remember to delete the install.packages() line or put a # in front of it 
```

### Opening Excel Dataset

- If the datafile is located in the same folder you can just type the name of the dataset in " "
- If the datafile is located in a different folder, you can find it by typing ../ and selecting enter
- For this assignment we will work on a STAT200 Dataset which examines test scores and variety of other predictors such as:
          - How much sleep students got
          - If students used computers or paper to write notes

Datasets can be in different formats. The most common forms are 'xlsx' and 'csv.'

```{r echo=TRUE, eval=FALSE}
# This is an example -- you will need to change the text inside paranthesis to the place where the data file is located ON YOUR OWN COMPUTER
# The most common format to run such functions are read_xlsx("file path in local computer")

Excel.dat.orginal <- read_xlsx("STAT200 Data.xlsx")
Excel.dat.orginal <- read_xlsx("../Stat 200 - Spring 2021/STAT200 Data.xlsx")

Excel.dat.working <- Excel.dat.orginal

Excel.dat.working
```

Opening CSV Dataset

```{r echo=TRUE,eval=FALSE}
CSV.dat.orginal <- read.csv("STAT200 Data.csv")
CSV.dat.orginal <- read.csv("../Stat 200 - Spring 2021/STAT200 Data.csv")

CSV.dat.working <- CSV.dat.orginal

CSV.dat.working
```

Make sure the excel & csv files match the function that is chosen! 
If you have an xlsx make sure to use the read_xlsx()

```{r echo=TRUE}
#error1 <- read_xlsx("STAT200 Data.csv")

```

## Basic Data set Functions

In this section we will show several basic dataset commands using the PSYC200 Dataset we loaded into R. Let us use the Excel File (Excel.dat). To learn more, watch [this](https://umd.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=6d0e4913-2524-4b55-8205-acd201718789&start=35.661219).

### Viewing Dataset

```{r echo=TRUE,eval=FALSE}
#View Entire Data
View(Excel.dat.working)

#Top 5 & Bottom 5
head(Excel.dat.working)
tail(Excel.dat.working)
```

### Structure & Summary of Dataset

```{r echo=TRUE,eval=FALSE}
#str() allows us to see the structure of the variables in the dataset
str(Excel.dat.working)

#summary provides us with a summary of the variables in the dataset
summary(Excel.dat.working)

```

### Descriptives of Dataset

```{r echo=TRUE,eval=FALSE}
#Number of Rows & Columns
nrow(Excel.dat.working)
ncol(Excel.dat.working)

#Column names
colnames(Excel.dat.working)

```


## More Dataset Functions

Here we explore more options to get some useful summaries from the data set. 

### Further Summaries of Dataset

```{r echo=TRUE,eval=FALSE}
#Located in install.packages() function
install.packages("psych")
library(psych)


##################
#Describe Function
##################
#describe() also provides us a clean summary/information on the variables in the dataset
describe(Excel.dat.working)
describe(Excel.dat.working, na.rm = TRUE)
  #describe(dataset, na.rm = TRUE/FALSE)
```

*Note: More information provided on describe function [here](https://www.rdocumentation.org/packages/psych/versions/2.0.12/topics/describe)*

`Describe By` does the same function as describe but it can separate the datasets out by groups...

- The PSYC200 dataset contains variables about GPA and variety of variables that affect GPA
- Ex: # Hours students sleep, Note Mode (paper notes or computer notes), etc.
    
```{r echo=TRUE, eval=FALSE}
#####################
#Describe By Function
#####################
#describe.by() provides us a clean summary/information on the variables in the dataset FOR A SPECIFIC GROUP

# Let's look at the means of test scores for each condition

describe.by(Excel.dat.working, Excel.dat.working$`Note Taking Modality (paper =1 or computer =2)`)
  #describe(dataset, dataset$group)
```

*Note: More information provided [here](https://www.rdocumentation.org/packages/psych/versions/2.0.12/topics/describeBy)*



