---
title: "Data Frame Manipulation and Exploratory Analysis"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "EDA"
  version: 0.5
---


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(psych)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

## Exploratory Data Analysis

###

In exploratory data analysis, we often begin by calculating summary statistics for and plotting key variables in our data. The `CityInfo` dataframe has both factor and numeric data. For example: it might be interesting to examine the relationship between population size and the number of sports teams or between the amount of sunshine and the number of bike commuters; or maybe we are interested in looking at regional variation in sunshine in our focal cities. 

Load in data and examine the structure.

```{r}

CityInfo  <- read.csv("D:/BSOS_R_Tutorial/data/Cityinfo.csv")
str(CityInfo)
```

Let's calculate the mean amount of `sunshine` for `midwestern` cities.

```{r include=TRUE}

mean(CityInfo[CityInfo$Region=="MW","Sunshine_per"])
```

Now let's subset the dataframe according to regions and calculate the mean sunshine for all regions. Try using the `aggregate()` function.

```{r include=TRUE}
aggregate(Sunshine_per~Region,CityInfo,mean)
```

## Practice Exercise-1

###

Often, we want to calculate numerous descriptive statistics simultaneously. There are numerous functions which can accomplish this task. Here, we'll try using functions from the `psych` package.


### Exercise 1.1

Using the `describe()` function, calculate summary statistics for the entire `CityInfo` dataframe.

```{r exercise1, exercise=TRUE}
```

```{r exercise1-check}
grade_result(
  pass_if(~ identical(.result, describe(CityInfo)), "Good Job!"),
  fail_if(~ TRUE)
)
```

### Exercise 1.2

Calculate summary statistics for only the number of sports teams per city.

```{r exercise2, exercise=TRUE}
```

```{r exercise2-check}
grade_result(
  pass_if(~ identical(.result, describe(CityInfo$Sportsteam_num)), "Good Job!"),
  fail_if(~ TRUE)
)
```

The `describeBy()` function is useful when you want to summarize numerous variables according to the some grouping variable.

### Exercise 1.3

Use `describeBy()` to generate summary statistics about the number of sports teams per city by region. Assign the results to a new object called `Sportstats`. Print the object `Sportstats`.


```{r exercise3, exercise=TRUE}
```

```{r exercise3-solution}
Sportstats <- describeBy(CityInfo$Sportsteam_num, CityInfo$Region)
Sportstats
```

```{r exercise3-check}
grade_code("Excellent!")
```


## Measures of Dispersion

###

In addition to central tendency metrics, it is also often useful to examine the spread or dispersion of observations within a dataset.

```{r}
# Make a histogram of the population data
hist(CityInfo$Populationmetro_2016)

# What is the variance in population?
var(CityInfo$Populationmetro_2016)

# you can also look at a range of summary statistics for populations
describe(CityInfo$Populationmetro_2016)
```


Another tool for quickly examining the spread of the data is the `table()` function. To answer the question `How many cities are in each region?` we can use the following code.

```{r}
table(CityInfo$Region)
```

## Practice Exercise-2


### Exercise 2_1

Find out the variation in amount of `sunshine` for the cities. 

```{r exercise4, exercise=TRUE}
```

```{r exercise4-check}
grade_result(
  pass_if(~ identical(.result, var(CityInfo$Sunshine_per)), "Good Job!"),
  fail_if(~ TRUE)
)
```


## Grouping Variables Together 

###

Oftentimes, data are not in the most useful format for particular analyses. We have to reshape and summarize data in order to draw insights from them. For this section, we will use data from: Tiffany Stephens and Ginny Eckert,2019, Knowledge Network for Biocomplexity. *urn:uuid:b910f74b-171b-4d2b-b065-fb21823a8e84*  
[Boat-based counts of sea otters at specific sites in Southeast Alaska](https://knb.ecoinformatics.org/view/urn%3Auuid%3Ab910f74b-171b-4d2b-b065-fb21823a8e84#urn%3Auuid%3A7eba259b-eeb5-4375-9596-e797bbb0b27d).
These data are sea otter counts from 2017 and 2018 in Alaska.

First read in and examine the data.
```{r}
otter <- read.csv("D:/BSOS_R_Tutorial/data/sea_otter_counts_2017&2018_CLEANDATA.csv")
str(otter)
```

This dataset has 8 variables: `region`, `site_name`, `latitude_N`, `latitude_E`, `data_DDMMYY`, `year`, `replicate`, and `n_otter`.

Let's aggregate the mean number of otters observed on a given day by site.

```{r eval=FALSE}
#data are aggregated based on site name and the function "mean" is applied to them
aggregate(otter, by=list(otter$site_name), FUN=mean, na.rm=TRUE)
```

The above applied the `mean()` function to all columns. This can be useful when you have many numeric variables you wish to quickly summarize. Other times however, it can be helpful to isolate specific variables for comparison.

```{r}
aggregate(n_otter~site_name, FUN=mean, data = otter)
```

## Practice Exercise-3

### Exercise 3.1

Using `order()` rearrange the dataframe by the total number of otters per site so that the sites with the most otters are at the top of the data. Store your resulting data frame in an object called `notterpersite`. Print it.

```{r exercise5, exercise=TRUE}
```

<div id="exercise5-hint">

**Hint** Try aggregating the data set first by site and then go ahead with `order()`.

</div>


```{r exercise5-solution}
notterpersite <- aggregate(formula=n_otter ~ site_name, FUN=sum, data=otter)
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
notterpersite
```

```{r exercise5-check}
grade_code("Excellent!")
```

### Exercise 3.2

What are the top 3 sites for otter sightings? Select out just these cases?

```{r exercise6, exercise=TRUE}
```

```{r exercise6-check}
grade_result(
  pass_if(~ identical(.result, notterpersite[1:3,]), "Good Job!"),
  fail_if(~ TRUE)
)
```



### Some More Exercises

Practice your data wrangling and exploratory data analysis skills by answering the following. Provide all code required to have only the required information as the output. 

### Exercise 1

Select the latitude and longitude of the site with the highest number of otter sightings on any single day.

```{r exercise7, exercise=TRUE}
```

```{r exercise7-check}
grade_result(
  pass_if(~ identical(.result, otter[otter$n_otter == max(otter$n_otter),c("latitude_N", "longitude_E")]), "Good Job!"),
  fail_if(~ TRUE)
)
```

### Exercise 2

How many observations come from Big Tree Day?

```{r exercise8, exercise = TRUE}
```

```{r exercise8-check}
grade_result(
  pass_if(~ identical(.result, length(otter$site_name[otter$site_name=="Big Tree Bay"])), "Good Job!"),
  fail_if(~ TRUE)
)
```

### Exercise 3

What was the total number of otters observed over the course of 2017?

```{r exercise9, exercise= TRUE}
```

```{r exercise9-check}
grade_result(
  pass_if(~ identical(.result, sum(otter[otter$year=="2017","n_otter"])), "Good Job!"),
  fail_if(~ TRUE)
)
```


## Using `ifelse()` statements

###

Another common form of logical testing in R is the `ifelse()` statement. In this case, you pass a logical test to R and if the output is true, a certain action is performed, then if it is false, another action is performed. This can be used to make new variables, subset data, color points on a graph and much more.

Using the `CityInfo` data and `ifelse()` we can mark each row based on whether or not the region is MW.

```{r}
ifelse(CityInfo$Region=="MW","yes","nope")
```

We can also see which cities are both in the midwest and have over 2% of commuters as bike commuters.

```{r}
ifelse(CityInfo$Region=="MW" & CityInfo$Bikecom_per>2,"yes","nope")
```

## Practice Exercise-4

`ifelse()` statements can also be nested. How might you write code to produce the output "sunny" for all cities with over 70 percent sunshine, and "kind of sunny" for cities with between 55 and 70 percent sunshine and "not sunny" for cities with less than 55 percent sunshine?

**Tip:** Add a new column to `CityInfo` dataframe named `sunlevel` to store the new information! Bind the `sunshine` amount and `sunlevel` columns together to display!

```{r exercise10, exercise=TRUE}
```

```{r exercise10-solution}
CityInfo$sunlevel<- ifelse(CityInfo$Sunshine_per > 70, "sunny",ifelse(CityInfo$Sunshine_per >55, "kind of sunny","not sunny"))
cbind(CityInfo$Sunshine_per,CityInfo$sunlevel)
```

```{r exercise10-check}
grade_code("Excellent!")
```


## The `paste()` function

###

Strings can be modified directly using the `paste()` function. This can be useful for creating new columns or systematically changing string data.

Fr example, we know that the `CityInfo$Year` column refers to the year each city was founded. Let's add the string "Founded in" before each year name.
 
```{r}
paste("Founded in", CityInfo$Year)
```

Now let's paste together the city and state names.

```{r}
paste(CityInfo$City,CityInfo$State)
```

An important feature of the `paste()` function is that you can customize the characters separating each value. 

```{r}

paste("Actor","Model","Writer", sep="-")

paste(CityInfo$City,CityInfo$State, sep=": ")

```


## Plotting in base R

###

A number of different types of charts can be made with the base graphics package.

For example,scatter plots can be made using the `plot()` function.

```{r}
plot(CityInfo$Populationmetro_2016,CityInfo$Sportsteam_num)
```

Line plots are made with the `plot()` function, while setting the type to line.

```{r}
plot(CityInfo$Populationmetro_2016,CityInfo$Sportsteam_num,type="l")
#This plot also conveys how R will sometimes give you an answer, even when the output is meaningless. It's up to you to distinguish meaningful results from nonsense.
```

Histograms can be made with the `hist()` function.

```{r}
hist(CityInfo$Bikecom_per)
```

Boxplots can be made with the `boxplot()` function.

```{r}
boxplot(Bikecom_per~Region,data=CityInfo)
```

## Adding labels

###

Adding descriptive labels is an important part of data visualization. In base R plotting, labels can be added with the `main=`, `xlab=` and `ylab=` arguments.

```{r}
boxplot(Bikecom_per~Region,data=CityInfo, main="Bike commuters by region", xlab="Region", ylab="Bike Commuters (%)")
```

Sometimes we also want to add annotations to plots noting things like sample size, key observations of interest, or summary statistics. Text can be added to an existing plot with the`text()` function. The position is set with x and y coordinates at the scale of the plot and the text can be customized in numerous ways.

```{r}
boxplot(Bikecom_per~Region,data=CityInfo, main="Bike commuters by region", xlab="Region", ylab="Bike Commuters (%)")
text(1,6,"Added text", col="blue")
```

Text can also be added to plots based on the position of data points in the plot. Here, we add text above each regional boxplot.


```{r}
boxplot(Bikecom_per~Region,data=CityInfo, main="Bike commuters by region", xlab="Region", ylab="Bike Commuters (%)")
text(CityInfo$Region,6,"Added text")
```

Add text over each boxplot at the mean point on each plot.

```{r}
boxplot(Bikecom_per~Region,data=CityInfo, main="Bike commuters by region", xlab="Region", ylab="Bike Commuters (%)")
text(1:4,aggregate(Bikecom_per ~ Region,FUN=mean, data=CityInfo)$Bikecom_per,"Added text", cex=0.7,col="blue")
```

## Practice Exercise-5

### Exercise 5.1

Label the bike commuter boxplot with the number of cities per region (that is, the n for each boxplot). Set the label up so that it will automatically update if the number of observations change. That is, find the number algorithmically rather than manually setting the number.

```{r exercise11, exercise=TRUE}
```

```{r exercise11-solution}
boxplot(Bikecom_per~Region,data=CityInfo, main="Bike commuters by region", xlab="Region", ylab="Bike Commuters (%)")
text(1:4,6,paste("N=",aggregate(City~Region,FUN=length, data=CityInfo)$City, sep=""))
```

```{r exercise11-check}
grade_code("Excellent!")
```


## Adjusting graphical parameters

###

By adjusting graphical parameters such as point shape, color, and size, you can represent multiple relationships in a single figure. In the figures below, we can differentiate the cities' regions using different shapes and color.

Below is a basic plot, with no customized graphical parameters.

```{r}
plot(CityInfo$Sunshine_per,CityInfo$Bikecom_per)
```

The `pch=` argument sets the shape of the points. In this case, we define the point shape based on the region each city point is in.
```{r}
plot(CityInfo$Sunshine_per,CityInfo$Bikecom_per,pch=c(CityInfo$Region))
```

The `col=` argument sets the color of the points. Note that the `Region` variable is a factor with 4 levels which is needed to be mentioned in order to group the points by different colors for different regions.

```{r}
plot(CityInfo$Sunshine_per,CityInfo$Bikecom_per,col=factor(CityInfo$Region))
```

The `cex=` argument sets the size of the points.
```{r}
plot(CityInfo$Sunshine_per,CityInfo$Bikecom_per,col=factor(CityInfo$Region),pch=19,cex=1.2)
```

## Practice Exercise-6

### Exercise 6.1

In the figures above, how do we know which symbol or color refers to which region? We have to make a legend. Try using the `legend()` to add a legend to your scatterplot. If you're stuck, take a look at the `legend()` help file.


```{r exercise12-practice, exercise=TRUE}
```

<div id="exercise12-practice-hint">
**Hint:**Create the plot first and then use `legend` on top of that. See at which position of the plot you want to display the legend!
</div>

### Pushing Things a Little Further!

The `Region` variable is a factor. This means it has unique properties we can use to our advantage., such as referring to the individual levels. It is also important to note that values are stored in the factor according to their level number, so if you want to convert the data to other classes, you must first adjust for this.

```{r}
plot(CityInfo$Sunshine_per,CityInfo$Bikecom_per,col=factor(CityInfo$Region),pch=19,cex=1.2)
legend(80,7, legend=levels(factor(CityInfo$Region)), fill=1:length(levels(factor(CityInfo$Region))), cex=0.7)
```


## Colors

###

R has 657 built-in colors. You can view them by using the `colors()` function. Try it!

```{r}
colors()
```

How did I know how many named colors there are? 

```{r}
length(colors())
```

At any given time, R will use a specific palette whenever colors are required. You can view this with the `palette()` function.

```{r}
palette()
```

This is the default palette. The colors are arranged sequentially, such that the first color used will be black, then red, then green3, etc. This is why in our plotting example above the colors could be referred to with their numerical index. 

The `RColorBrewer` package makes beautiful palettes for data visualization in R. Check out the palettes here: [ColorBrewer2.0](http://colorbrewer2.org/).

We can also explore a few more useful palettes.
```{r fig.height=7, fig.width= 4}
rpal<-rainbow(4)
hpal<-heat.colors(4)
tpal<-topo.colors(4)


par(mfrow=c(2,2))
plot(CityInfo$Populationmetro_2016,CityInfo$Bikecom_per,col=factor(CityInfo$Region),pch=19,cex=1.2)
plot(CityInfo$Populationmetro_2016,CityInfo$Bikecom_per,col=rpal[factor(CityInfo$Region)],pch=19,cex=1.2)
plot(CityInfo$Populationmetro_2016,CityInfo$Bikecom_per,col=hpal[factor(CityInfo$Region)],pch=19,cex=1.2)
plot(CityInfo$Populationmetro_2016,CityInfo$Bikecom_per,col=tpal[factor(CityInfo$Region)],pch=19,cex=1.2)

```

## Symbol shapes

###

There are 25 different shapes available in base R. Some can have different colors for the border and fill, try them out as you continue exploring plotting in R. Read more [here](http://www.statmethods.net/advgraphs/parameters.html)

## Exploring data with graphs

###

The `pairs()` function quickly shows the correlations between all variables in a dataframe. This can reveal potential combinations of variables that warrant further investigation.

```{r}
pairs(CityInfo[4:8])
```




