---
title: "Data sets in R"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "R-vectors"
  version: 0.5
---

```{r setup, include=FALSE}
library(MASS)
library(learnr)
library(gradethis)
library(readxl)
library(xlsx)
library(datasets)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
learnr::tutorial_options(
  exercise.timelimit = 60)
#load("D:/BSOS_R_Tutorial/PSYC_Tutorials/data_psych/exam_scores (1).rda")
#write.xlsx(exam_scores,"D:/DATA FOR PSYCH/STAT200_Data.xlsx")
```


## Section One : Extracting Rows and Columns

In this module, we will learn some basic data set operations.

- We will start this lesson by using the class dataset we used in Lesson 3
- Let's start by using the csv class data file

### Opening/loading Data set

```{r echo=TRUE}
dat.org <- read.csv("D:/BSOS_R_Tutorial/PSYC_Tutorials/data_psych/STAT200_Data_csv.csv")
dat.work <- dat.org

#Note: Make sure you are using readxl instead of readx1
#You will need to redo this part to open the dataset in your folder
```

### Revision of Related Functions

In this section we will talk about a few exploratory functions that can help you explore the data set more. Carefully look at the following code chunk to understand what each function does.

```{r echo=TRUE}
#Name of the columns
colnames(dat.work) #See all the variable names in dataset

#General Functions
str(dat.work) #See structure of all variables in dataset
summary(dat.work) #See numerical/character/factor summary of all variables in dataset

#Head/Tail
head(dat.work) #see first few rows
tail(dat.work) #see last few rows

#View
View(dat.work) #View entire dataset
```


### Extracting Specific Variables (columns)

In this section we will see how to extract a specific variable from the entire dataset. Often we will need to look at one single variable or do some mathematical operations on some specific variables.

```{r echo=TRUE}
#This will help us visualize what we will do... 
head(dat.work)

#This is how to Extract a specific variable via $
dat.work$sleep_week
dat.work$motivation

#This is how to Extract a specific variables via [ ]
dat.work[,2] # extracting the values for only column 2
dat.work[,3] # extracting the values for only column 3
dat.work[,2:3] #extracting the values for only column 2 and 3
```

*Note:* While using [ ] for extracting variables, the format should be [Row, Column]. 


### Extracting Specific Cases(Rows)

```{r echo=TRUE}
#Help Visualize what we will do... 
head(dat.work)

#Extracting specific rows (Row 1, Row 5 & Row 1:5)
dat.work[1,]
dat.work[5,]
dat.work[1:5,]
```

### Selecting Specific Variables & rows in the Dataset

```{r echo=TRUE}
#Help Visualize what we will do... 
head(dat.work)

#Here we can see that a [1,] indicates row 1 & [,1] indicates column 1
dat.work[1,]
dat.work[,1]


#Extracting specific variables & rows

#Here we extract row 1 and column 1
dat.work[1,1]

#Here we extract row 1 to 3 & column 1 to 3
dat.work[1:3, 1:3]
```

### Try it!

1. Load the dataset "airquality" from the package "datasets" and store it as an object called *airquality*. Print the first few rows of the dataset.

```{r code_sandbox, exercise=TRUE, exercise.cap="exercise1"}
```

2. Choose the correct one.

```{r question1}
question("To extract the solar radiation amounts in the air, which of the following ways can be used?",
         answer("airquality$Solar.R", correct=TRUE, message = "Good job!"),
         answer("airquality$2", correct = FALSE),
         answer("airquality[2,]",  correct = FALSE),
         allow_retry = TRUE,
         random_answer_order = TRUE
)

learnr::question("How can you extract the values of first 3 variables for the first 3 cases?",
         answer("airquality[1:3,1:3]", correct=TRUE, message = "Good job!"),
         answer("airquality[1:3,]", correct = FALSE),
         answer("airquality[,1:3]",  correct = FALSE),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```


## Section Two : Extracting & Building Vectors / Dataframes

### Creating Vectors from Variables in Datasets

```{r echo=TRUE}
#Help Visualize what we will do... 
head(dat.work)

#Now we access Student Sleep variable in the dataset
Sleep <- dat.work$sleep_week
Sleep

#Next we access Student Motivation in the dataset
Motivation <- dat.work$motivation
Motivation

#Now we access student Exam Scores in the dataset
Scores <- dat.work$exam_score
Scores
```

### Selecting specific Parts of Vectors

We can also use the [row,column] function for specific values in a vector. Note that we can only pull the rows here since this is a vector and **NOT** a dataframe.


```{r echo=TRUE}
#First 3 Rows in Sleep
Sleep[1:3]

#First 3 Rows in Motivation
Motivation[1:3]

#First 3 Rows in Scores
Scores[1:3]
```

### Creating New Vectors 

Here we will do some basic arithmetic using RStudio, except now we will reference specific variables in the dataset using the $ .
  
```{r echo=TRUE}
#New Vectors
#Creating Vectors from the Dataset (and doing Arithmetic)

Sleep.DividedbyTwo <- dat.work$sleep_week/2

Motivation.MultipliedbyTwo <- dat.work$motivation*2

Scores.Squared <- dat.work$exam_score^2

Combining.Vectors <- dat.work$sleep_week + dat.work$motivation + dat.work$exam_score

# Viewing New Vectors
head(Sleep.DividedbyTwo)
head(Motivation.MultipliedbyTwo)
head(Scores.Squared)
head(Combining.Vectors)
```

### Alternative Take on Creating New Vectors

We can create these same vectors by pulling them directly out of the dataset. Here we do this; and then select only the first 3 rows/rows of data.
    
```{r echo=TRUE}
#New Vectors
#Creating Vectors from the Dataset (first 3 rows/rows only)

First3.Sleep <- dat.work$sleep_week[1:3]
First3.Motivation <- dat.work$motivation[1:3]
First3.Scores <- dat.work$exam_score[1:3]

#Viewing New Vectors
First3.Sleep
First3.Motivation
First3.Scores
```

Now try doing some arithmetic by your own!

### Try it!

Use the pre-loaded data set on airquality for this exercises.

1. Create a vector that subtracts 78 from each of the values of the variable `Temp`. Divide this new vector by 9. Store the resulting vector as `std_temp`.

```{r q3, exercise = TRUE}
```

```{r q3-check}
grade_result(
  pass_if(~ identical(.result, (airquality$Temp-78)/9), "Good Job!"),
  fail_if(~ TRUE)
)
```


## Section Three : Subsetting

In this section we will now "subset" specific variables and values in RStudio.


### Review Original Dataset Structure & Review Factors

```{r echo=TRUE}
#Review Structure of Original Data
str(dat.org)

#Change Statistics Anxiety Variable from Numeric to Factor
dat.work$anxiety <- as.factor(dat.work$anxiety)
    #1 - Low Anxiety
    #2 - Medium Anxiety
    #3 - High Anxiety


#Checking if the change worked!
str(dat.org$anxiety) #Original --> Numeric
str(dat.work$anxiety) #Working --> Factor

levels(dat.work$anxiety) #This will show us each of the factors of the variable
```

### Factor Subsetting

We briefly discussed Factors last week with the Describe.by() function. For details, refer back to [this](https://umd.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=772f3370-5d2e-4fb4-8e23-acd201762d34&start=405.655013)

```{r echo=TRUE}
library(psych)
describe.by(dat.work, dat.work$anxiety)

    #Here we see 3 datasets which reflect a summary of the variables anxiety Factor
          #1 - Low Anxiety #Mean exam score = 80.33
          #2 - Medium Anxiety #Mean exam score = 89.73
          #3 - High Anxiety #Mean exam score = 76.43

```

### Specific method to Subset Variables by Factors

Here we look at exam scores by each factor of *anxiety* by directly referencing the factor we are interested in using [ ] and then ==
You can see this below.
    
```{r echo=TRUE}
#Subset factors directly using the [] function

#Exam Scores for Students with Low Anxiety (anxiety == 1)
dat.work$exam_score[dat.work$anxiety == 1]

#Exam Scores for Students with medium Anxiety (anxiety == 2)
dat.work$exam_score[dat.work$anxiety == 2]

#Exam Scores for Students with High Anxiety (anxiety == 3)
dat.work$exam_score[dat.work$anxiety == 3]

#Mean of Scores per Each Subsetted Factor

#Mean of Exam Scores for Students with Low Anxiety (anxiety == 1)
mean(dat.work$exam_score[dat.work$anxiety == 1])

#Mean of Exam Scores for Students with Medium Anxiety (anxiety == 2)
mean(dat.work$exam_score[dat.work$anxiety == 2])

#Mean of Exam Scores for Students with High Anxiety (anxiety == 3)
mean(dat.work$exam_score[dat.work$anxiety == 3])

  #Note:Compare these means with the ones provided in the describe.by() function!
        #1 - Low Anxiety #Mean = 80.33
        #2 - Medium Anxiety #Mean = 89.73
        #3 - High Anxiety #Mean = 76.43

```

We can use these subsets when we examine correlations for different groups! This is very useful when we do data analysis.Here we see the correlation between exam Scores & the amount of sleep a week before exam."

```{r echo=TRUE}
#Correlation Test between Sleep & Test Scores

cor1 <- cor.test(dat.work$exam_score, dat.work$sleep_week)
cor1

#Correlation Test between Sleep & Test Scores for people who had Low anxiety
cor2 <- cor.test(dat.work$exam_score[dat.work$anxiety == 1], dat.work$sleep_week[dat.work$anxiety == 1])
cor2

#Correlation Test between Sleep & Test Scores for people who had High anxiety
cor3 <- cor.test(dat.work$exam_score[dat.work$anxiety == 3], dat.work$sleep_week[dat.work$anxiety == 3])
cor3

###################
#Note: anxiety
###################
  #1 = Low Anxiety
  #2 = Medium Anxiety
  #3 = High Anxiety

```

### Other Logical Operators

There are other logical operators for subsetting which are briefly reviewed below.

### Equal to/Not Equal to

```{r echo=TRUE}
## Equal to
#Equal to 1 (Low Anxiety)
dat.work$exam_score[dat.work$anxiety == 1]

mean(dat.work$exam_score[dat.work$anxiety == 1])

## Not Equal to
#Not Equal to 1 (Not low Anxiety)
dat.work$exam_score[dat.work$anxiety != 1]

mean(dat.work$exam_score[dat.work$anxiety != 1])
```

### AND

```{r echo=TRUE}
#Preparing Dataset
#First Need Converting Note Taking Type at Time Period 1
dat.work$Note_Type_T1 <- as.factor(dat.work$note_mode)

#Viewing to See if it Worked
str(dat.work$Note_Type_T1)
levels(dat.work$Note_Type_T1)
    #Paper = 1
    #Computer = 2

#AND LOGICAL OPERATOR
#&

#Exam Scores for Low Anxiety (1) & Note Taking Mode Paper (1)
dat.work$exam_score[dat.work$anxiety == 1 & dat.work$Note_Type_T1 == 1]

#The Mean
mean(dat.work$exam_score[dat.work$anxiety == 1 & dat.work$Note_Type_T1 == 1])

#Length (How many rows qualified)
length(dat.work$exam_score[dat.work$anxiety == 1 & dat.work$Note_Type_T1 == 1])

    
#Exam Scores for High Anxiety (3) & Note Taking Mode Computer (2)
dat.work$exam_score[dat.work$anxiety == 3 & dat.work$Note_Type_T1 == 2]

#The Mean
mean(dat.work$exam_score[dat.work$anxiety == 3 & dat.work$Note_Type_T1 == 2])

#Length (How many rows qualified)
length(dat.work$exam_score[dat.work$anxiety == 3 & dat.work$Note_Type_T1 == 2])
```

There are many other logical operators also!

### Less than / Less than or equal to Logical Operator

```{r echo=TRUE}
#Less than (<)
dat.work$exam_score[dat.work$sleep_week < 2520]

#The Mean    
mean(dat.work$exam_score[dat.work$sleep_week < 2520])
    
#Length (How many rows qualified)
length(dat.work$exam_score[dat.work$sleep_week < 2520])

#Less than or equal to (<=)
dat.work$exam_score[dat.work$sleep_week <= 2520]
    
#The Mean    
mean(dat.work$exam_score[dat.work$sleep_week <= 2520])
    
#Length (How many rows qualified)
length(dat.work$exam_score[dat.work$sleep_week<= 2520])
```


### Greater than / greater than or equal to Logical Operator

```{r echo=TRUE}
#Greater than (>)
dat.work$exam_score[dat.work$sleep_week > 2520]

#The Mean    
mean(dat.work$exam_score[dat.work$sleep_week > 2520])
    
#Length (How many rows qualified)
length(dat.work$exam_score[dat.work$sleep_week > 2520])

#Greater than or equal to (>=)
dat.work$exam_score[dat.work$sleep_week >= 2520]
    
#The Mean    
mean(dat.work$exam_score[dat.work$sleep_week >= 2520])
    
#Length (How many rows qualified)
length(dat.work$exam_score[dat.work$sleep_week >= 2520])
```

Try some subsetting by your own!

### Try it!

For this set of exercises, use `airquality` again.

1. Convert the `Month` variable into a factor. 

```{r q4, exercise = TRUE}
```

```{r q4-check}
grade_result(
  pass_if(~ identical(.result, as.factor(airquality$Month)), "Good Job!"),
  fail_if(~ TRUE)
)
```

2. What are the observations for wind speed in month 6?

```{r q5, exercise=TRUE}
```

```{r q5-check}
grade_result(
  pass_if(~ identical(.result, airquality$Wind[airquality$Month==6]), "Good Job!"),
  fail_if(~ TRUE)
)
```

3. Using subsetting on `airquality`, print the wind speed values for month 6 and `Temp` being less than or equal to 80.

```{r q6, exercise=TRUE}
```

```{r q6-check}
grade_result(
  pass_if(~ identical(.result, airquality$Wind[airquality$Month==6 & airquality$Temp <=80]), "Good Job!"),
  fail_if(~ TRUE)
)
```

4. Answer the following.

```{r question7}
learnr::question("Select the correct way to check the number of observations you had after the last subsetting operation, i.e. in question 3.",
         answer("length(airquality$Month==6 & airquality$Temp <=80)", correct=FALSE),
         answer("length(airquality$Wind[airquality$Month==6 & airquality$Temp <=80])", correct = TRUE,  message = "Good job!"),
         answer("length(airquality$Wind[airquality$Month=6 & Temp <=80])",  correct = FALSE),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

If you want to find more on these operations, check the following links!
[operators](https://www.statmethods.net/management/operators.html)

[subset](https://www.statmethods.net/management/subset.html)



